HydroFlow: A New Dataflow Programming Language

### 2022-10-31 paper.tex
a forwarding pointer

@@

In this work we present the HydroFlow programming language, a new dataflow language and runtime designed to overcome these three limitations.

\section{Background: The three major problems of existing streaming frameworks}

@@

When streaming computations are paired with state (as is their common case), thorny issues of consistency and determinism can quickly arise.  Consider, for example, an operator that runs a single query φ on the entire stream as a single unit, producing a single boolean result ($φ : stream → bool$). Implementing such operators are challenging; existing systems generally take one of three approaches.  Industry-standard dataflow environments (like Spark and such) take the conservative approach; they wait for a distinguished end-of-stream message\MPM{say something intelligent about batching here or something}, after which they are able to safely run φ on the entire stream.  Other systems, most notable differential dataflow, take a different approach; they will continuously re-evaluate φ on each input, issuing a _retraction_ if the result of φ should ever change.  It is then the responsibility of downstream operators to determine how to use this retraction to adjust their own computational results.  Within the PL space, reactive programming frameworks have solved this problem in a number of ways; most commonly, upon the arrival of an updated value, they re-evaluate all functions which depended on that value, using a memoization strategy to avoid performing truly redundant computation.

@@

Meanwhile, both retractions and re-executions feature significant wasted computation.  In the best case, retractions must effectively produce _new_ computation to undo the effects of old, resulting in many CPU-hours wasted; in the worst case, retractions arrive too late, opening the application up to consistency issues.  Re-executions, meanwhile, are unable to determine ahead of time if the re-execution will meaningfully alter the value ultimately computed by the reactive program, stopping only when a cached value has been dynamically detected.  This can lead to significant wasted re-computation; for example, when computing if the cardinality of a set is above a certain threshold, these computations would continuously re-compute the set cardinality, only to determine that it has no effect on the threshold and thus no effect on the overall computation.

@@

>>>>

If you're looking for the working document with all my mathy notes about the type system, head over to "notes/types.tex"

@@

In this work we present the HydroFlow programming language, a new dataflow language and runtime designed to overcome these three limitations. I should say a sentence here about the basics of HydroFlow, like users write Operators in a traditional systems programming language, and combine operators into stream-processing programs via a simple graph description syntax. First, HydroFlow's strong static type system is capable of encoding not just the types of data elements that flow through streams, but also _properties_ of the streams themselves: for example, the type of items in a hydroflow stream might statically indicate that stream elements are comparable, while the _stream properties_ may ensure that these elements arrive in _monotonically ascending_ order.  This strong type system can statically detect many more type errors, preventing run-time crashes. Second, HydroFlow can leverage its stream properties to perform a _monotonicity analysis_, indicating whether a given operator is safe to run _incrementally_, and thereby solving the latency issues encountered by the introduction of barriers in traditional stream processing systems, or the consistency issues which arise from the unwise elision thereof. Finally, HydroFlow addresses dataflow scalability concerns by introducing rewriting-based optimizations familiar to the world of SQL directly into the dataflow language itself. In HydroFlow, users may provide new operators along with an _axiomatic specification_ of their behavior; the HydroFlow compiler than uses an SMT-based analysis to determine potential rewrite axioms that relate all operators, allowing a traditional SQL-style optimization approach to full HydroFlow programs---even in the presence of arbitrary user-provided code.

Hydroflow is implemented as a deep embedding into the Rust programming language \cite{rust}.  Our compilation stack ensures all operators are compiled alongside their assembled graph, allowing the Rust compiler to perform aggressive inlining and copy-elision, erasing the overhead of  operator handoffs and greatly improving on the performance of similar tools such as timely dataflow \cite{timely-dataflow}.  In particular, our tool demonstrates some quantifiable improvement in some benchmark or other that will prove quite compelling here.

The rest of the paper proceeds as follows.  Section \ref{sec:background}
 provides background on the challenges faced by existing distributed dataflow environments. Section \ref{sec:hydroflow-overview} provides an overview of the hydroflow programming language, with examples of both hydroflow programs and implementing operators. Section \ref{sec:hydroflow-semantics} describes the formal semantics of hydroflow, with a focus on the semantics of stream properties. Section \ref{sec:hydroflow-types} describes the HydroFlow type system, Section \ref{sec:hydroflow-implementation} describes the implementation, and \ref{sec:experimental-results} reviews its performance on several benchmarks.  We also have a related work and conclusion somewhere.

\section{Background: The three major problems of existing streaming frameworks}
\label{sec:background}

@@

When streaming computations are paired with state (as is their common case), thorny issues of consistency and determinism can quickly arise.  Consider, for example, an operator that runs a single query φ on the entire stream, producing a single boolean result ($φ : stream → bool$). Implementing such operators is challenging; existing systems generally take one of three approaches.  Industry-standard dataflow environments (like Spark and such) take the conservative approach; they wait for a distinguished end-of-stream message\MPM{say something intelligent about batching here or something}, after which they are able to safely run φ on the entire stream.  Other systems, most notably differential dataflow, take a different approach; they will continuously re-evaluate φ on each input, issuing a _retraction_ if the result of φ should ever change.  It is then the responsibility of downstream operators to determine how to use this retraction to adjust their own computational results.  Within the PL space, reactive programming frameworks have solved this problem in a number of ways; most commonly, upon the arrival of an updated value, a reactive pipeline will be re-run up until a previously-computed intermediate result is encountered.

@@

Meanwhile, both retractions and re-executions feature significant wasted computation.  In the best case, retractions must effectively produce _new_ computation to undo the effects of old, resulting in many CPU-hours wasted; in the worst case, retractions arrive too late, opening the application up to consistency issues.  Re-executions, meanwhile, are unable to determine ahead of time if the re-execution will meaningfully alter the value computed by the reactive program, stopping only if a cached intermediate result has been dynamically detected.\MPM{I feel like I need to read more FRP to make sure this is actually true, and I shouldn't be re-pitching this as "bringing an FRP insight to dataflow" or something.}  This can lead to significant wasted re-computation; for example, when computing if the cardinality of a set is above a certain threshold, these computations would continuously re-compute the set cardinality, only to determine that it has no effect on the threshold and thus no effect on the overall computation.

@@

Distributed frameworks are often lauded for their scalability; indeed, the ability to schedule various operators across and entire clusters of nodes, rather than have fixed roles for each machine, produces a degree of dynamic flexibility and allows architectures to scale the resources provided to each individual operator.  But these scaling approaches can only go so far; they rely on finding different execution strategies for the same graph, but ignore opportunities for optimization which require rewriting the graph itself.

SQL optimizers, meanwhile, have long relied on the ability to rewrite dataflow graphs. A SQL query engine generates a dataflow graph from a SQL query, and can reason equationally about SQL queries in order to discover a distinct dataflow program which is capable of more efficiently computing the desired result. While this often produces a strictly-more-optimal query, SQL engines (unlike traditional compiler optimizations) have the benefit of knowing something about the data on which the program will be running; they can thus make optimization decisions which are tailored to the expected data, and thus can achieve far greater efficiency.

Bringing these benefits to the general domain of distributed dataflow programming poses a challenge: SQL optimizers assume complete knowledge of program text, and allow only a small set of operators.  Crucially, there are no abstractions in these programs beyond tables and no extensibility or modularity beyond ``user-defined functions (UDFs),'' which must be treated conservatively by the optimizer and which allow only a restrictive class of additional computations.  In contrast, the world of distributed dataflow programming frequently features custom operators, and increasingly relies on abstractions in the form of both libraries and modular program design \cite{hope-this-is-true!}. Thus, modern dataflow libraries largely make no attempt to perform rewriting optimizations; those which do perform rewriting optimizations simply defer to SQL as an input language, allowing traditional query optimization assumptions to hold.
